---
title: 'Data Preperation for Bird Species in Comparative Analysis'
author: "Justin G. Cally"
subtitle: Supplementary Material
output:
  pdf_document: default
  html_document:
    toc: true # table of content true
    toc_float: false # make 
    depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: false  ## if you want number sections at each table header
    theme: yeti # lovely fonts and colours
    code_folding: hide # awesome buttons to show/hide the code
---

#Bird Environmental Data

Here we will try and extract environmental data for a polygon that estimates a species distribution. 

Load in raster and sp package
```{r, warning=FALSE, message=FALSE}
library(raster)
library(sp)
#devtools::install_github("r-spatial/sf") #install gdal via brew
library(sf)
library(rgdal)
library(maptools)
library(pander)
library(plyr)
library(dplyr)
library(tidyverse)
library(knitr)
library(lsr)
library(data.table)
library(readxl)
library(fuzzyjoin)
library(parallel)
# devtools::install_github("jimhester/archive") # Need to use: brew install libarchive in bash
# devtools::install_github("mirzacengic/climatedata")
library(climatedata)
options(stringsAsFactors = FALSE)
```

##Read in all species

Because the HBW taxonomy changes year to year we will have to use a field known as the SpcRecID and the SISRecID. SpcRecID was used pre-2014 and thus is used for the birdtree.org phylogeny. SISRecID is used post 2014 and is thus used for this edition of the spatial dataset. 

```{r}
if(!file.exists("data/Intermediate_data/BOTW.rds")){
all.data <- sf::read_sf("data/BOTW/BOTW.gdb") #read_sf is supposed to be faster than readOGR
saveRDS(all.data, "data/Intermediate_data/BOTW.rds")
} else {
  all.data <- readRDS("data/Intermediate_data/BOTW.rds")
  }
nest.data <- read_csv("data/Combined_allspp.csv", ) %>% mutate(SCINAME = str_replace_all(species, "_", " "))

# HBW checklists from 2010 (when Jetz made phylogeny) and most recent for shapefiles
HBW.checklist.2010 <- read_excel("taxa_details/BirdLife_Checklist_Version_3.xls") %>% select(`Scientific name`, contains("Synonyms"), contains("SpcRecID")) %>% `colnames<-`(c("SCINAME", "Synonyms", "SpcRecID"))
HBW.checklist.2018 <- read_excel("taxa_details/HBW-BirdLife_Checklist_Version_3.xlsx", skip = 1) %>% select(`Scientific name`, contains("SpcRecID"), contains("SISRecID")) %>% `colnames<-`(c("SCINAME", "SpcRecID", "SISRecID"))

# Read in the metadata for the Jetz phylogeny. We should use these names in the phylogenetic analysis
Jetz.Species <- read.csv("taxa_details/Jetz_Species.csv") %>% select(WJSpecID, Scientific)

# Read in the data that had non-matching taxanomic names/missing species ID
missing.species.id <- read.csv("taxa_details/no_SpcRecID_IM.csv") %>% 
  mutate(SCINAME = str_replace_all(species, "_", " ")) %>% 
  select(SpcRecID.new = SpcRecID, SCINAME) %>% na.omit()
jetz.taxonomy.missing <- read.csv("taxa_details/Jetz_taxonomy_missing.csv")
```


```{r missingdata}
#### Data match to phylogeny ####
# Group SpecRecID as a vector column as some species have multiple IDs. 
HBW.checklist.2010.collapsed <- HBW.checklist.2010 %>%
  filter(!is.na(SCINAME))
  # group_by(SCINAME) %>%
  # summarise(SpcRecID = I(list(SpcRecID)))
# sum(is.null(HBW.checklist.2010.collapsed$SpcRecID))
# # Using the nest data add the Species IDs (matching by SCINAME and also synonyms)
nest.data.withID <- nest.data %>% left_join(HBW.checklist.2010.collapsed)
# length(which(sapply(nest.data.withID$SpcRecID, is.null)))
# # 56
# # Match by synonyms
nest.data.withID2 <- HBW.checklist.2010 %>% select(-SCINAME, Syn.SpcRecID = SpcRecID) %>%
  fuzzy_right_join(nest.data.withID, by = c("Synonyms" = "SCINAME"), match_fun = str_detect) %>% 
  mutate(SpcRecID = coalesce(SpcRecID, Syn.SpcRecID))
  # mutate(SpcRecID = case_when(!is.na(Syn.SpcRecID) ~ pmap(list(SpcRecID, Syn.SpcRecID), c),
  #                             is.na(Syn.SpcRecID) ~ pmap(list(SpcRecID), c)))
# length(which(sapply(nest.data.withID2$SpcRecID, is.null)))
# # 44

# # Add in the missing species ~ 40 species 
nest.data.withID3 <- nest.data.withID2 %>% left_join(missing.species.id) %>%
  mutate(SpcRecID = coalesce(SpcRecID.new, as.integer(SpcRecID)))
#   mutate(SpcRecID = case_when(!is.na(SpcRecID.new) ~ pmap(list(SpcRecID, SpcRecID.new), c),
#                               is.na(SpcRecID.new) ~ pmap(list(SpcRecID), c)))
# length(which(sapply(nest.data.withID3$SpcRecID, is.null)))
# # 12 (removed)

nest.data.jetz <- nest.data.withID3 %>% 
  left_join(jetz.taxonomy.missing %>% select(SCINAME, sp_tax_jetz)) %>%
  mutate(SCINAME = coalesce(sp_tax_jetz, SCINAME)) %>%
  left_join(Jetz.Species, by = c("SCINAME" = "Scientific")) %>%
  filter(!is.na(WJSpecID)) # REMOVE 3 SPECIES NOT INCLUDED IN JETZ

#### Data match to shapefiles ####
species.with.shapes <- all.data %>% as.data.frame() %>% select(SISID) %>% unique()

#### Species to extract data for ####
nest.2018.ID.data <- nest.data.jetz %>% 
  inner_join(HBW.checklist.2018 %>% rename(SCINAME2018 = SCINAME), by = "SpcRecID") %>%
  filter(!is.na(SpcRecID)) %>%
  group_by(SISRecID) %>%
  summarise(SpcRecID = list(unique(SpcRecID)),
            SCINAME2018 = list(unique(SCINAME2018)),
            Jetz.name = list(unique(SCINAME))) %>%
  ungroup()

if(!file.exists("data/Intermediate_data/bird.ranges.breeding.Rds")){
bird.ranges <- all.data %>% 
  filter(SISID %in% nest.2018.ID.data$SISRecID & SEASONAL %in% c(1,2)) %>% #Resident or breeding ranges
  st_cast("MULTIPOLYGON") %>% #Remove MULTISURFACE shapes that ruin everything
  group_by(SISID) %>%
  summarise(SCINAME = glue::glue_collapse(SCINAME, sep = ", "), 
            SEASONAL = glue::glue_collapse(SEASONAL, sep = ", "),
            Shape = st_union(Shape), #combine the multiple polygons for each species
            Range.Size = st_area(Shape))
saveRDS(bird.ranges, file = "data/Intermediate_data/bird.ranges.breeding.Rds")
} else {
  bird.ranges <- readRDS("data/Intermediate_data/bird.ranges.breeding.Rds")
}

```

Run loop
```{r, warning = FALSE}
# #Only use the breeding ranges
# for (i in 1:length(nest.2018.ID.data)) {
#   tmp <- nest.2018.ID.data[nest.2018.ID.data$SCINAME == unique[i], ] 
#   writeOGR(tmp, dsn='data/All_species_R', unique[i], driver="ESRI Shapefile",
#            overwrite_layer=TRUE)
#    }
```

Great, now we have all our individual shapefiles named as ``Abeillia abeillei.shp`` 

##Worldclim data

Now extract worldclim data: we use grids of 2.5 min
```{r}
### Get Data Chelsea for the current period
chelsa_bioclim_current <- get_chelsa(type = "bioclim",
                                     output_dir = "data/climate",
                                     period = "current") #Downloads first time but then reads in tifs

#It's now saved in the data folder

### Get Future climate predictions: Run this later

# From these models we want to narrow it down to a subset that has low interdependency; this means we get good estimates of uncertainty without having to extract data from all the models. This ranking is taken from: Sanderson, B.M., Knutti, R. & Caldwell, P. (2015) A Representative Democracy to Reduce Interdependency in a Multimodel Ensemble. Journal of Climate, 28, 5171–5194)
top5.low.interdependent.models <- c("CESM-1-BCG",
                                  "MPI-ESM-MR",
                                  "MIROC5",
                                  "CMCC-CM",
                                  "CESM1-CAM5")


# Get models with all 4 RCP scenarios
models_all_rcp <- check_models() %>% 
group_by(model) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n == 4) %>%
  distinct(model) %>%
  pull()

if(!dir.exists("data/climate/current")){
chelsa_bioclim_future <- get_chelsa(output_dir = my_output_directory, 
                                      period = "future", 
                                      future_years = "2041-2060", 
                                      scenario_string = "rcp85", #Get multiple and interpolate
                                      model_string = models_all_rcp)
}
```

#Now we want to randomly sample each polygon
```{r, warning=FALSE, message=FALSE, results='hide'}
if(!file.exists("data/intermediate_data/bird.points.rds")){
#Increse the iterations (defaukt is 4) so we can obtain complete samples of each range
bird.range.list <- setNames(split(bird.ranges, 
                                  seq(nrow(bird.ranges))), 
                            bird.ranges$SISID)
bird.points <- list()
bird.points <- mclapply(bird.range.list, function(x){
  sp::spsample(sf::as_Spatial(x), n = 1000, type = "random", iter = 30, quiet = T)
}, mc.cores = 4)
saveRDS(bird.points, "data/intermediate_data/bird.points.rds")
} else {
bird.points <- readRDS("data/intermediate_data/bird.points.rds")
}
```

Now we extract the bioclim data from each of the points
```{r}
bird.values <- mclapply(bird.points, function(x) {
  raster::extract(chelsa_bioclim_current, x)
  }, mc.cores = 4) #raster cant deal with sf 
#bird.values <- as.list(data.frame((bird.values)))
```

From these 1000 points we want a mean and sd. We can do this when building the df

```{r}

bird.frame <- mclapply(bird.values, function(x) {as.data.frame(x)}, mc.cores = 4)
bird.summary <- mclapply(bird.frame, function(x) {
  as.data.table(x)[, c(mean = lapply(.SD, mean, na.rm=TRUE), 
                       sd = lapply(.SD, sd, na.rm=TRUE)),]
}, mc.cores = 4)
bird.summary.combined <- cbind(data.table(SISID = names(bird.points)),
                                    do.call(rbind, bird.summary))
saveRDS(bird.summary.combined, "data/Intermediate_data/bird.summary.combined.rds")

#Combine with original data 
bird.data.recombined <- bird.summary.combined %>% 
  as.data.frame() %>%
  mutate(SISID = as.integer(SISID)) %>%
  left_join(bird.ranges %>% 
              st_drop_geometry() %>%
              select(SISID, SCINAME, Range.Size) %>%
              as.data.frame())

saveRDS(bird.data.recombined, "data/Intermediate_data/bird.data.recombined.rds")
```




transpose
```{r}
bird.means <- t(bird.summary)
bird.means <- split(bird.means, 1:19)
```

Now add column of species name
```{r}
bird.means <- cbind.data.frame(shps.jetz, bird.means)
```

```{r}
rownames(bird.means) <- NULL
colnames(bird.means) <- c("binomial","bioclim1", "bioclim2", "bioclim3", "bioclim4", "bioclim5", "bioclim6", "bioclim7", "bioclim8", "bioclim9", "bioclim10", "bioclim11", "bioclim12", "bioclim13", "bioclim14", "bioclim15", "bioclim16", "bioclim17", "bioclim18", "bioclim19")
```

Write csv
```{r}
write.csv(bird.means, 'data/bird.means.csv')
```

##Range size

Cool! Ok, let's try and get range sizes from the list.
```{r}
bird.range.size <- sapply(bird.ranges,
                       function(x) {(area(x))})
bird.range.size <- sapply(bird.range.size, function(x) {sum(x)})
bird.range.size <- as.data.frame(bird.range.size)
bird.range.size <- cbind.data.frame(shps.jetz, bird.range.size)
rownames(bird.range.size) <- NULL
colnames(bird.range.size) <- c("binomial", "range.size.m2")
```

Write csv
```{r}
write.csv(bird.range.size, 'data/bird.range.size.csv')
```






##Now for Paleoclimate

Using the CCSM4 models we download 2.5 minute raster grids of predicted climate for the last glacial maximum (22,000 years ago). We then need to import this GeoTiff into the environment and into a rasterstack

```{r}
bi.layer <- dir('data/LGMdata', "*.tif")
LGM.stack <- stack(paste(getwd(),"/data/LGMdata/", bi.layer, sep=""))
```

Since we already have 1000 samples of each bird range we can proceed to sampling the paleoclimate layers via the following: 

Now we extract the bioclim data from each of the points
```{r}
LGM.bird.values <- lapply(bird.points, function(x) {raster::extract(LGM.stack, x)})


LGM.bird.values <- as.list(data.frame((LGM.bird.values)))
```

From these 1000 points we want a min, mean, max. We can do this when building the df

```{r}

LGM.bird.frame <- lapply(LGM.bird.values, function(x) {as.data.frame(x)})
LGM.bird.summary <- llply(LGM.bird.frame, function(x) {
  (as.data.frame(apply(x,2,mean, na.rm =T)))})
```

transpose
```{r}
LGM.bird.means <- t(as.data.frame(LGM.bird.summary))
LGM.bird.means <- (split(LGM.bird.means, 1:19))

#Divide temperatures by 10
LGM.bird.means[c(1,2,3,12,14:19)] <- llply(LGM.bird.means[c(1,2,3,12,14:19)], function(x) {
  ifelse(is.na(x),x,x/10)})
```

Now add column of species name
```{r}
LGM.bird.means <- cbind.data.frame(shps.jetz, LGM.bird.means)
```

Get rid of rownames
```{r}
rownames(LGM.bird.means) <- NULL
colnames(LGM.bird.means) <- c("binomial","cclgmbi1", "cclgmbi10", "cclgmbi11", "cclgmbi12", "cclgmbi13", "cclgmbi14", "cclgmbi15", "cclgmbi16", "cclgmbi17", "cclgmbi18", "cclgmbi19", "cclgmbi2", "cclgmbi3", "cclgmbi4", "cclgmbi5", "cclgmbi6", "cclgmbi7", "cclgmbi8", "cclgmbi9") #automatically arranges in strange order
```

Write csv
```{r}
write.csv(LGM.bird.means, 'data/LGM.bird.means.csv')
```

Alongside LGM we can also get the last inter-glacial (~120,000 - 140,000 BP)

```{r}
LIG.layer <- dir('data/LIGdata', "*.bil")
LIG.stack <- stack(paste(getwd(),"/data/LIGdata/", LIG.layer, sep=""))
```

Since we already have 1000 samples of each bird range we can proceed to sampling the paleoclimate layers via the following: 

Now we extract the bioclim data from each of the points
```{r}
LIG.bird.values <- lapply(bird.points, function(x) {raster::extract(LIG.stack, x)})


LIG.bird.values <- as.list(data.frame((LIG.bird.values)))
```

From these 1000 points we want a min, mean, max. We can do this when building the df

```{r}
LIG.bird.frame <- lapply(LIG.bird.values, function(x) {as.data.frame(x)})
LIG.bird.summary <- llply(LIG.bird.frame, function(x) {
  (as.data.frame(apply(x,2,mean, na.rm =T)))})
```

transpose
```{r}
LIG.bird.means <- t(as.data.frame(LIG.bird.summary))
LIG.bird.means <- (split(LIG.bird.means, 1:19))

#Divide temperatures by 10
LIG.bird.means[c(1,2,3,12,14:19)] <- llply(LIG.bird.means[c(1,2,3,12,14:19)], function(x) {
  ifelse(is.na(x),x,x/10)})

```

Now add column of species name
```{r}
LIG.bird.means <- cbind.data.frame(shps.jetz, LIG.bird.means)
```

```{r}
rownames(LIG.bird.means) <- NULL
colnames(LIG.bird.means) <- c("binomial","LIG.bi1", "LIG.bi10", "LIG.bi11", "LIG.bi12", "LIG.bi13", "LIG.bi14", "LIG.bi15", "LIG.bi16", "LIG.bi17", "LIG.bi18", "LIG.bi19", "LIG.bi2", "LIG.bi3", "LIG.bi4", "LIG.bi5", "LIG.bi6", "LIG.bi7", "LIG.bi8", "LIG.bi9")
```

Write csv
```{r}
write.csv(LIG.bird.means, 'data/LIG.bird.means.csv')
```

##NPP

We obtain 30 arc-sec estimates of annual NPP through the Moderate Resolution Imaging Spectroradiometer (MODIS picture). Specifically the Numerical Terradynamic Simulation Group (NTSG) has developed a database with algorithms that take into account corrections etc. to produce a GeoTIFF of mean NPP from 2000 to 2015. Using Each Species we can extract the NPP... how do we do this?

http://files.ntsg.umt.edu/


```{r}
NPP <- raster('data/NPP/NPP00_15.tif')
```

Now we extract the NPP data from each of the points
```{r}
NPP.bird.values <- lapply(bird.points, function(x) {extract(NPP, x)})
```

From these 1000 points we want a min, mean, max. We can do this when building the df

```{r}
NPP.bird.frame <- lapply(NPP.bird.values, function(x) {as.data.frame(x)})
NPP.bird.summary <- lapply(NPP.bird.frame, function(x) {
  (as.data.frame(apply(x,2,mean, na.rm =T)))})
```

transpose
```{r}
NPP.bird.means <- t(as.data.frame(NPP.bird.summary))
```

Now add column of species name
```{r}
NPP.bird.means <- cbind.data.frame(shps.jetz, NPP.bird.means)
```

Get rid of rownames
```{r}
rownames(NPP.bird.means) <- NULL
colnames(NPP.bird.means) <- c("binomial", "NPP")
```

Write csv
```{r}
write.csv(NPP.bird.means, 'data/NPP.bird.means.csv')
```


##Calculate the variance

We want to calculate the variance in our sampling. This can be done from taking ``var()`` for the dataframes in both present and past climate.

###For present climate
```{r}
bird.se.summary <- lapply(bird.frame, function(x) {
  (as.data.frame(apply(x,2,function(x) sqrt(var(x,na.rm = TRUE)/length(na.omit(x))))))})

#transpose
bird.se <- t(as.data.frame(bird.se.summary))

#split
bird.se <- as.data.frame(split(bird.se, 1:19))

#Now add column of species name
bird.se <- cbind.data.frame(shps.jetz, bird.se)

#remove rownames
rownames(bird.se) <- NULL
colnames(bird.se) <- c("binomial","se.bioclim1", "se.bioclim2", "se.bioclim3", "se.bioclim4", "se.bioclim5", "se.bioclim6", "se.bioclim7", "se.bioclim8", "se.bioclim9", "se.bioclim10", "se.bioclim11", "se.bioclim12", "se.bioclim13", "se.bioclim14", "se.bioclim15", "se.bioclim16", "se.bioclim17", "se.bioclim18", "se.bioclim19")

#Write csv
write.csv(bird.se, 'data/bird.se.csv')
```



Standard Deviation
```{r}
bird.sd.summary <- lapply(bird.frame, function(x) {
  (as.data.frame(apply(x,2,function(x) sd(x,na.rm = TRUE))))})

#transpose
bird.sd <- t(as.data.frame(bird.sd.summary))

#split
bird.sd <- as.data.frame(split(bird.sd, 1:19))

#Now add column of species name
bird.sd <- cbind.data.frame(shps.jetz, bird.sd)

#remove rownames
rownames(bird.sd) <- NULL
colnames(bird.sd) <- c("binomial","sd.bioclim1", "sd.bioclim2", "sd.bioclim3", "sd.bioclim4", "sd.bioclim5", "sd.bioclim6", "sd.bioclim7", "sd.bioclim8", "sd.bioclim9", "sd.bioclim10", "sd.bioclim11", "sd.bioclim12", "sd.bioclim13", "sd.bioclim14", "sd.bioclim15", "sd.bioclim16", "sd.bioclim17", "sd.bioclim18", "sd.bioclim19")

#Write csv
write.csv(bird.sd, 'data/bird.sd.csv')
```


```{r}
library(ggplot2)
sd.or.se <- left_join(bird.se, bird.sd)
cor(sd.or.se$se.bioclim1, sd.or.se$sd.bioclim1)

sd.or.se %>% ggplot(aes(x = se.bioclim3, y = sd.bioclim3))+ geom_point()

cor(x = sd.or.se$se.bioclim1, y = sd.or.se$sd.bioclim1)
```


###For LGM

Run the same code for the LGM dataset
```{r}
LGM.bird.sd.summary <- lapply(LGM.bird.frame, function(x) {
  (as.data.frame(apply(x,2,function(x) sd(x,na.rm = TRUE))))})

#transpose
LGM.bird.sd <- t(as.data.frame(LGM.bird.sd.summary))

#split
LGM.bird.sd <- as.data.frame(split(LGM.bird.sd, 1:19))

#Divide temperatures by 10 
LGM.bird.sd[c(1,2,3,12,14:19)] <- LGM.bird.sd[c(1,2,3,12,14:19)]/10

#Now add column of species name
LGM.bird.sd <- cbind.data.frame(shps.jetz, LGM.bird.sd)

#remove rownames
rownames(LGM.bird.sd) <- NULL
colnames(LGM.bird.sd) <- c("binomial","sd.cclgmbi1", "sd.cclgmbi10", "sd.cclgmbi11", "sd.cclgmbi12", "sd.cclgmbi13", "sd.cclgmbi14", "sd.cclgmbi15", "sd.cclgmbi16", "sd.cclgmbi17", "sd.cclgmbi18", "sd.cclgmbi19", "sd.cclgmbi2", "sd.cclgmbi3", "sd.cclgmbi4", "sd.cclgmbi5", "sd.cclgmbi6", "sd.cclgmbi7", "sd.cclgmbi8", "sd.cclgmbi9") #automatically arranges in strange order

#Write csv
write.csv(LGM.bird.sd, 'data/LGM.bird.sd.csv')
```

###For NPP

```{r}
NPP.bird.sd.summary <- lapply(NPP.bird.frame, function(x) {
  (as.data.frame(apply(x,2,function(x) sd(x,na.rm = TRUE))))})

#transpose
NPP.bird.sd <- t(as.data.frame(NPP.bird.sd.summary))

#Now add column of species name
NPP.bird.sd <- cbind.data.frame(shps.jetz, NPP.bird.sd)

#remove rownames
rownames(NPP.bird.sd) <- NULL
colnames(NPP.bird.sd) <- c("binomial", "NPP.sd")

#Write csv
write.csv(NPP.bird.sd, 'data/NPP.bird.sd.csv')
```

###For LIG


```{r}
LIG.bird.sd.summary <- lapply(LIG.bird.frame, function(x) {
  (as.data.frame(apply(x,2,function(x) sqrt(var(x,na.rm = TRUE)/length(na.omit(x))))))})

#transpose
LIG.bird.sd <- t(as.data.frame(LIG.bird.sd.summary))

#split
LIG.bird.sd <- as.data.frame(split(LIG.bird.sd, 1:19))

#Divide temperatures by 10 
LIG.bird.sd[c(1,2,3,12,14:19)] <- LIG.bird.sd[c(1,2,3,12,14:19)]/10

#Now add column of species name
LIG.bird.sd <- cbind.data.frame(shps.jetz, LIG.bird.sd)

#remove rownames
rownames(LIG.bird.sd) <- NULL
colnames(LIG.bird.sd) <- c("binomial","sd.LIG.bi1", "sd.LIG.bi10", "sd.LIG.bi11", "sd.LIG.bi12", "sd.LIG.bi13", "sd.LIG.bi14", "sd.LIG.bi15", "sd.LIG.bi16", "sd.LIG.bi17", "sd.LIG.bi18", "sd.LIG.bi19", "sd.LIG.bi2", "sd.LIG.bi3", "sd.LIG.bi4", "sd.LIG.bi5", "sd.LIG.bi6", "sd.LIG.bi7", "sd.LIG.bi8", "sd.LIG.bi9")

#Write csv
write.csv(LIG.bird.sd, 'data/LIG.bird.sd.csv')
```


###Combining the data: 
In excel we joined the data together and double checked the values

##Assess correlations

We may want to discard predictors as although they are loaded into a PCA, having highly correlated predictors may overestimate loadings and/or variance. Additionally they will increase computational requirements. We will be using the following function to assess correlations: 

```{r}
#use correlation assessment function
panel.cor <- function(x, y, digits=2, prefix="", cex.cor, ...)
{
usr <- par("usr"); on.exit(par(usr))
par(usr = c(0, 1, 0, 1))
r <- abs(cor(x, y, use="pairwise.complete.obs"))
txt <- format(c(r, 0.123456789), digits=digits)[1]
txt <- paste0(prefix, txt)
if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
text(0.5, 0.5, txt, cex = cex.cor * r)
}
```

**The bioclimate variables assessed here are:**

BIO1 = Annual Mean Temperature

BIO2 = Mean Diurnal Range (Mean of monthly (max temp - min temp))

BIO3 = Isothermality (BIO2/BIO7) (* 100)

BIO4 = Temperature Seasonality (standard deviation *100)

BIO5 = Max Temperature of Warmest Month

BIO6 = Min Temperature of Coldest Month

BIO7 = Temperature Annual Range (BIO5-BIO6)

BIO8 = Mean Temperature of Wettest Quarter

BIO9 = Mean Temperature of Driest Quarter

BIO10 = Mean Temperature of Warmest Quarter

BIO11 = Mean Temperature of Coldest Quarter

BIO12 = Annual Precipitation

BIO13 = Precipitation of Wettest Month

BIO14 = Precipitation of Driest Month

BIO15 = Precipitation Seasonality (Coefficient of Variation)

BIO16 = Precipitation of Wettest Quarter

BIO17 = Precipitation of Driest Quarter

BIO18 = Precipitation of Warmest Quarter

BIO19 = Precipitation of Coldest Quarter


###Current day bioclim variables


```{r, fig.width=15, fig.height=15}
#Draw plot
pairs(bird.means[,2:20], lower.panel=panel.smooth, upper.panel=panel.cor)
```

###For LGM data

```{r, fig.width=15, fig.height=15}
#Draw plot
pairs(LGM.bird.means[,2:20], lower.panel=panel.smooth, upper.panel=panel.cor)
```


###Human inteference data

Night lights

Distance to major cities

Population density


We can obtain population density gridded estimates from the following: 

Center for International Earth Science Information Network - CIESIN - Columbia University. 2017. Gridded Population of the World, Version 4 (GPWv4): Population Density Adjusted to Match 2015 Revision UN WPP Country Totals, Revision 10. Palisades, NY: NASA Socioeconomic Data and Applications Center (SEDAC). https://doi.org/10.7927/H49884ZR. Accessed DAY MONTH YEAR

The following is 30 arc seconds of adjusted density from 2010

```{r}
pop.density <- raster("data/human_impact/pop_density_2010_30arc.tif")
```

Now we extract the NPP data from each of the points
```{r}
pop.density.values <- lapply(bird.points, function(x) {raster::extract(pop.density, x)})
```

From these 1000 points we want a min, mean, max. We can do this when building the df

```{r}
pop.density.frame <- lapply(pop.density.values, function(x) {as.data.frame(x)})
pop.density.summary <- lapply(pop.density.frame, function(x) {
  (as.data.frame(apply(x,2,mean, na.rm =T)))})
```

transpose
```{r}
pop.density.means <- t(as.data.frame(pop.density.summary))
```

Now add column of species name
```{r}
pop.density.means <- cbind.data.frame(shps.jetz, pop.density.means)
```

Get rid of rownames
```{r}
rownames(pop.density.means) <- NULL
colnames(pop.density.means) <- c("binomial", "Pop_Density")
```

Write csv
```{r}
write.csv(pop.density.means, 'data/Pop.density.means.csv')
```


```{r}
pop.density.sd.summary <- lapply(pop.density.frame, function(x) {
  (as.data.frame(apply(x,2,function(x) sqrt(var(x,na.rm = TRUE)/length(na.omit(x))))))})

#transposd
pop.density.sd <- t(as.data.frame(pop.density.sd.summary))

#Now add column of species name
pop.density.sd <- cbind.data.frame(shps.jetz, pop.density.sd)

#remove rownames
rownames(pop.density.sd) <- NULL
colnames(pop.density.sd) <- c("binomial", "pop.density.sd")

#Write csv
write.csv(pop.density.sd, 'data/pop.density.sd.csv')

```

##Join data together


```{r}
#Masterlist from Jetz/birdtree master
prelim.dataframe <- read.csv('data/Pass_Master_Tax.csv')

#Bioclim.means
#bioclim.means <- read.csv('data/bird.means.csv')
prelim.dataframe <- left_join(prelim.dataframe, bird.means, by = 'binomial')

#Bioclim.sd
#bioclim.sd <- read.csv('data/bird.sd.csv')
prelim.dataframe <- left_join(prelim.dataframe, bird.sd, by = 'binomial')

#Range Size
prelim.dataframe <- left_join(prelim.dataframe, bird.range.size, by = 'binomial')

#LGM means
#LGM.means <- read.csv('data/LGM.bird.means.csv')
prelim.dataframe <- left_join(prelim.dataframe, LGM.bird.means, by = 'binomial')

#LGM sd
#LGM.sd <- read.csv('data/LGM.bird.sd.csv')
prelim.dataframe <- left_join(prelim.dataframe, LGM.bird.sd, by = 'binomial')

#LIG means
#LIG.means <- read.csv('data/LIG.bird.means.csv')
prelim.dataframe <- left_join(prelim.dataframe, LIG.bird.means, by = 'binomial')

#LIG sd
#LIG.sd <- read.csv('data/LIG.bird.sd.csv')
prelim.dataframe <- left_join(prelim.dataframe, LIG.bird.sd, by = 'binomial')

#NPP means
#NPP.means <- read.csv('data/NPP.bird.means.csv')
prelim.dataframe <- left_join(prelim.dataframe, NPP.bird.means, by = 'binomial')

#NPP sd
#NPP.sd <- read.csv('data/NPP.bird.sd.csv')
prelim.dataframe <- left_join(prelim.dataframe, NPP.bird.sd, by = 'binomial')

#Pop Density means
#Pop.density.means <- read.csv('data/Pop.density.means.csv')
prelim.dataframe <- left_join(prelim.dataframe, pop.density.means, by = 'binomial')

#Pop density sd
#Pop.density.sd <- read.csv('data/Pop.density.sd.csv')
prelim.dataframe <- left_join(prelim.dataframe, pop.density.sd, by = 'binomial')

write.csv(prelim.dataframe, 'data/complete.dataframe2.csv', row.names = F)
```


<!-- ##Obtaining an LGM anomaly for temperature and precipitation -->

<!-- First for mean annual temperature BIO1 -->
<!-- ```{r} -->
<!-- #For LGM  -->
<!-- LGM.BIO1 <- raster('data/LGMdata/cclgmbi1.tif') -->

<!-- #Extract annual temp values from raster -->
<!-- LGM.BIO1.values <- lapply(bird.points, function(x) {extract(LGM.BIO1, x)}) -->

<!-- ``` -->


<!-- And now annual mean temp for current climate -->
<!-- ```{r} -->
<!-- #Subset raster from stack -->
<!-- BIO1.current <- bioclim_data@layers[[1]] -->

<!-- #Extract annual temp values from raster -->
<!-- BIO1.values.current <- lapply(bird.points, function(x) {extract(BIO1.current, x)}) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- BIO1.df <- as.data.frame(BIO1.values.current) -->
<!-- BIO1.combined <- c(BIO1.values.current, LGM.BIO1.values) -->
<!-- diff <- mapply(function(X,Y) { -->
<!--   t.test(X, Y) -->
<!--   }, X=BIO1.values.current, Y=LGM.BIO1.values) -->

<!-- ``` -->

##R Session Info

```{r}
sessionInfo() %>% pander
```

